---
title: "Simplify Functional Enrichment Results"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output: 
  rmarkdown::html_vignette:
    fig_caption: true
    css: main.css
vignette: >
  %\VignetteIndexEntry{1. Simplify Functional Enrichment Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center",
    dev = "jpeg"
)
options(width = 100)
```

The **simplifyEnrichment** package clusters functional terms into groups by
clustering the similarity matrix of the terms with a new proposed method
"binary cut" which recursively applies partition around medoids (PAM) with two
groups on the similarity matrix and in each iteration step, a score is
assigned to decide whether the group of gene sets that corresponds to the
current sub-matrix should be split or not. For more details of the method, 
please refer to the simplifyEnrichment paper.

## Simplify GO enrichment results

```{r, echo = FALSE, message = FALSE}
library(simplifyEnrichment)
mat = readRDS(system.file("extdata", "random_GO_BP_sim_mat.rds", package = "simplifyEnrichment"))
go_id = rownames(mat)
```

The major use case for **simplifyEnrichment** is for simplying the GO
enrichment results by clustering the corresponding semantic similarity matrix
of the significant GO terms. To demonstrate the usage, we first generate a
list of random GO IDs from the Biological Process (BP) ontology category:

```{r, eval = FALSE}
library(simplifyEnrichment)
set.seed(888)
go_id = random_GO(500)
```

**simplifyEnrichment** starts with the GO similarity matrix. Users can use
their own similarity matrices or use the `GO_similarity()` function to
calculate the semantic similarity matrix. The `GO_similarity()` function is
simply a wrapper on `GOSemSim::termSim()`. The function accepts a vector of GO
IDs. Note the GO terms should only belong to one same ontology (_i.e._, `BP`,
`CC` or `MF`).


```{r, eval = FALSE}
mat = GO_similarity(go_id)
```

By default, `GO_similarity()` uses `Rel` method in `GOSemSim::termSim()`. Other
methods to calculate GO similarities can be set by `measure` argument, _e.g._:

```{r, eval = FALSE}
GO_similarity(go_id, measure = "Wang")
```

With the similarity matrix `mat`, users can directly apply `simplifyGO()`
function to perform the clustering as well as visualizing the results.

```{r, fig.width = 7*1.2, fig.height = 4*1.2}
df = simplifyGO(mat)
```

On the right side of the heatmap there are the word cloud annotations which
summarize the functions with keywords in every GO cluster. Note there is no
word cloud for the cluster that is merged from small clusters (size < 5).

The returned variable `df` is a data frame with GO IDs, GO terms and the
cluster labels:

```{r}
head(df)
```

The size of GO clusters can be retrieved by:

```{r}
sort(table(df$cluster))
```

Or split the data frame by the cluster labels:

```{r, eval = FALSE}
split(df, df$cluster)
```

`plot` argument can be set to `FALSE` in `simplifyGO()`, so that no plot is
generated and only the data frame is returned.

If the aim is only to cluster GO terms, `binary_cut()` or `cluster_terms()` functions can be
directly applied:

```{r}
binary_cut(mat)
```

or

```{r, eval = FALSE}
cluster_terms(mat, method = "binary_cut")
```

`binary_cut()` and `cluster_terms()` basically generate the same clusterings, but the labels of clusters might differ.

## Simplify general functional enrichment results

Semantic measurements can be used for the similarity of GO terms. However,
there are still a lot of ontologies (e.g. MsigDB gene sets) that are only
represented as a list of genes where the similarity between gene sets are
mainly measured by gene overlap. **simplifyEnrichment** provides the
`term_similarity()` and other related functions
(`term_similarity_from_enrichResult()`, `term_similarity_from_KEGG()`,
`term_similarity_from_Reactome()`, `term_similarity_from_MSigDB()` and
`term_similarity_from_gmt()`) which calculate the similarity of terms by the
gene overlapping, with methods of [Jaccard
coefficient](https://en.wikipedia.org/wiki/Jaccard_index), [Dice
coefficient](https://en.wikipedia.org/wiki/S%C3%B8rensen%E2%80%93Dice_coefficient),
[overlap coefficient](https://en.wikipedia.org/wiki/Overlap_coefficient) and
[kappa coefficient](https://en.wikipedia.org/wiki/Cohen%27s_kappa).

The similarity can be calculated by providing:

1. A list of gene sets where each gene set contains a vector of genes.
2. A `enrichResult` object which is normally from the 'clusterProfiler', 'DOSE', 'meshes' or 'ReactomePA' package.
3. A list of KEGG/Reactome/MsigDB IDs. The gene set names can also be provided for MsigDB ontologies.
4. A gmt file and the corresponding gene set IDs.

Once you have the similarity matrix, you can send it to `simplifyEnrichment()` function.
But note, as we benchmarked in the manuscript, the clustering on the gene
overlap similarity performs much worse than on the semantic similarity.

## Comparing clustering methods

In the **simplifyEnrichment** package, there are also functions that compare
clustering results from different methods. Here we still use previously
generated variable `mat` which is the similarity matrix from the 500 random GO
terms. Simply running `compare_clustering_methods()` function performs all supported
methods (in `all_clustering_methods()`) excluding `mclust`, because
`mclust` usually takes very long time to run. The function generates a figure
with three panels:

1. A heatmap of the similarity matrix with different clusterings as row
   annotations.
2. A heatmap of the pair-wise concordance of the clustering from every two methods.
3. Barplots of the difference scores for each method, the number of clusters
   (total clusters and the clusters with size >= 5) and the mean similarity of
   the terms that are in the same clusters (block mean).

In the barplots, the three metrics are defined as follows:

1. **Different score**: This is the difference between the similarity values
   for the terms that belong to the same clusters and different clusters.
   For a similarity matrix $M$, for term $i$ and term $j$ where $i \ne j$, the
   similarity value $x_{i,j}$ is saved to the vector $\mathbf{x_1}$ only when
   term $i$ and $j$ are in a same cluster. $x_{i,j}$ is saved to the vector
   $\mathbf{x_2}$ when term $i$ and $j$ are not in the same cluster. The
   difference score measures the distribution difference between $\mathbf{x_1}$
   and $\mathbf{x_2}$, calculated as the Kolmogorov-Smirnov statistic between
   the two distributions.
2. **Number of clusters**: For each clustering, there are two numbers: the
   number of total clusters and the number of clusters with size >= 5 (only
   the big clusters).
3. **Block mean**: The mean similarity values of the blocks in the similarity
   heatmap. Similar denotation as difference score, for term $i$ and $j$ where
   $i$ can be the same as $j$ (values on the diagonal are also used), the
   similarity value $x_{i,j}$ is saved to the vector $\mathbf{x_3}$ only when
   term $i$ and $j$ are in a same cluster. The block mean is the mean value
   over $\mathbf{x_3}$.

```{r, fig.width = 10, fig.height = 7}
compare_clustering_methods(mat)
```

If `plot_type` argument is set to `heatmap`. There are heatmaps for the
similarity matrix under different clusterings methods. The last panel is a
table with the number of clusters.

```{r, fig.width = 18, fig.height = 14, dev = "jpeg"}
compare_clustering_methods(mat, plot_type = "heatmap")
```

Please note, the clustering methods might have randomness, which means,
different runs of `compare_clustering_methods()` may generate different clusterings
(slightly different). Thus, if users want to compare the plots between
`compare_clustering_methods(mat)` and `compare_clustering_methods(mat, plot_type = "heatmap")`, they
should set the same random seed before executing the function.

```{r, eval = FALSE}
set.seed(123)
compare_clustering_methods(mat)
set.seed(123)
compare_clustering_methods(mat, plot_type = "heatmap")
```

`compare_clustering_methods()` is simply a wrapper on `cmp_make_clusters()`
and `cmp_make_plot()` functions where the former function performs
clustering with different methods and the latter visualizes the results. To
compare different plots, users can also use the following code without
specifying the random seed.

```{r, eval = FALSE}
clt = cmp_make_clusters(mat) # just a list of cluster labels
cmp_make_plot(mat, clt)
cmp_make_plot(mat, clt, plot_type = "heatmap")
```

### Register new clustering methods

New clustering methods can be added by `register_clustering_methods()`,
removed by `remove_clustering_methods()` and reset to the default methods by
`reset_clustering_methods()`. All the supported methods can be retrieved by
`all_clustering_methods()`. `compare_clustering_methods()` runs all the clustering methods
in `all_clustering_methods()`.

The new clustering methods should be as user-defined functions and sent to
`register_clustering_methods()` as named arguments, e.g.:

```{r, eval = FALSE}
register_clustering_methods(
    method1 = function(mat, ...) ...,
    method2 = function(mat, ...) ...,
    ...
)
```

The functions should accept at least one argument which is the input matrix
(`mat` in above example). The second optional argument should always be `...`
so that parameters for the clustering function can be passed by `control`
argument from `cluster_terms()` or `simplifyGO()`. If users forget to add
`...`, it is added internally.

Please note, the user-defined function should automatically identify the
optimized number of clusters. The function should return a vector of cluster
labels. Internally it is converted to numeric labels.

## Examples

There are following examples which we did for the benchmarking in the manuscript:

- [Examples of simplifyEnrichment](https://simplifyenrichment.github.io/examples/).
- [Compare different similarity measurements for functional terms](https://simplifyenrichment.github.io/compare_similarity/).
- [Compare different partitioning methods in binary cut clustering](https://simplifyenrichment.github.io/test_partition_methods/).


## Session Info

```{r}
sessionInfo()
```
